---
title: "SDU Nest Box Project"
title-block: false
date-modified: last-modified
execute:
  freeze: auto
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(scales)

# Paths into the sibling SDUBirdProject repo
sdu_root <- file.path(dirname(getwd()), "SDUBirdProject")
data_processed <- function(...) file.path(sdu_root, "Data", "processed", ...)
data_raw       <- function(...) file.path(sdu_root, "Data", "raw", ...)

all_data <- read_csv(data_processed("nest_checks", "allData.csv"),
                     show_col_types = FALSE) |>
  mutate(
    Timestamp      = as.POSIXct(Timestamp, tz = "Europe/Copenhagen"),
    nEggs_num      = suppressWarnings(as.numeric(na_if(as.character(nEggs), "AON"))),
    nLiveChicks_num = suppressWarnings(as.numeric(na_if(as.character(nLiveChicks), "AON"))),
    year   = Year,
    doy    = Day,
    box_id = as.character(boxNumber)
  )

gps_path <- data_raw("gps", "SDUNestboxLocation.csv")
if (file.exists(gps_path)) {
  box_locations <- read_csv(gps_path, show_col_types = FALSE) |>
    transmute(
      box_number = suppressWarnings(as.integer(Name)),
      longitude  = Longitude,
      latitude   = Latitude
    )
} else {
  box_locations <- tibble(box_number = integer(),
                          longitude  = numeric(),
                          latitude   = numeric())
}

years_monitored <- length(unique(all_data$year))
total_visits    <- nrow(all_data)
total_box_ids   <- n_distinct(all_data$box_id)

nest_year_summary <- all_data |>
  group_by(year, box_id) |>
  summarise(
    max_eggs   = if (all(is.na(nEggs_num)))      NA_real_ else max(nEggs_num,      na.rm = TRUE),
    max_chicks = if (all(is.na(nLiveChicks_num))) NA_real_ else max(nLiveChicks_num, na.rm = TRUE),
    species    = dplyr::last(na.omit(species)),
    .groups    = "drop"
  ) |>
  mutate(
    occupied       = !is.na(max_eggs)   & max_eggs   > 0,
    chicks_present = !is.na(max_chicks) & max_chicks > 0
  )

total_occupied_nests <- sum(nest_year_summary$occupied, na.rm = TRUE)
median_clutch_size   <- median(nest_year_summary$max_eggs[nest_year_summary$occupied], na.rm = TRUE)

occupancy_by_year <- nest_year_summary |>
  group_by(year) |>
  summarise(
    occupied_nests    = sum(occupied,       na.rm = TRUE),
    nests_with_chicks = sum(chicks_present, na.rm = TRUE),
    .groups = "drop"
  )

first_egg <- all_data |>
  filter(!is.na(nEggs_num) & nEggs_num > 0) |>
  group_by(year, box_id) |>
  filter(doy == min(doy)) |>
  ungroup() |>
  mutate(estimated_first_egg_day = doy - nEggs_num)

first_egg_by_year <- first_egg |>
  group_by(year) |>
  summarise(
    median_first_egg_day = median(estimated_first_egg_day, na.rm = TRUE),
    q25 = quantile(estimated_first_egg_day, 0.25, na.rm = TRUE),
    q75 = quantile(estimated_first_egg_day, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

species_lookup <- c(
  BT = "Blue Tit", GT = "Great Tit", NH = "Nuthatch",
  MT = "Marsh Tit", WR = "Wren", `Other (make a note!)` = "Other"
)

species_summary <- nest_year_summary |>
  filter(occupied, !is.na(species), species != "") |>
  mutate(species_public = recode(species, !!!species_lookup, .default = species)) |>
  count(species_public, sort = TRUE) |>
  mutate(prop = n / sum(n))

study_area_summary <- nest_year_summary |>
  mutate(box_number = suppressWarnings(as.integer(sub("_.*", "", box_id)))) |>
  distinct(box_number) |>
  inner_join(box_locations, by = "box_number")

api_path <- data_raw("nestboxes_io", "nestbox_record_all_records.csv")
has_api  <- file.exists(api_path)

if (has_api) {
  api_data <- read_csv(api_path, show_col_types = FALSE) |>
    mutate(
      record_date     = as.Date(record_date),
      record_datetime = as.POSIXct(record_datetime, tz = "Europe/Copenhagen"),
      year   = lubridate::year(record_date),
      box_id = as.character(box_id),
      eggs   = suppressWarnings(as.numeric(eggs)),
      chicks = suppressWarnings(as.numeric(chicks))
    )
  current_year <- max(api_data$year, na.rm = TRUE)
  current_data <- api_data |> filter(year == current_year)
  current_summary <- current_data |>
    summarise(
      checks          = n(),
      boxes_checked   = n_distinct(box_id),
      boxes_with_eggs = n_distinct(box_id[!is.na(eggs)   & eggs   > 0]),
      boxes_with_chicks = n_distinct(box_id[!is.na(chicks) & chicks > 0]),
      latest_date     = max(record_date, na.rm = TRUE)
    )
  checks_by_day <- current_data |>
    count(record_date, name = "n_checks") |>
    arrange(record_date)
} else {
  current_year <- NA_integer_
}
```

```{=html}
<style>
.hero {
  background: linear-gradient(135deg, #1f4f40 0%, #8e5a2b 100%);
  color: #fbfaf6;
  padding: 34px 36px;
  border-radius: 26px;
  margin: 10px 0 26px 0;
  box-shadow: 0 16px 36px rgba(31, 79, 64, 0.18);
}
.hero h1 { color: #fbfaf6; margin: 0 0 8px 0; font-size: 2.4rem; }
.hero p  { max-width: 820px; margin: 0; font-size: 1.06rem; color: rgba(251,250,246,0.9); }

.stat-grid {
  display: grid;
  grid-template-columns: repeat(5, minmax(0, 1fr));
  gap: 14px;
  margin: 22px 0 24px 0;
}
.stat-card {
  background: rgba(255,255,255,0.8);
  border: 1px solid rgba(31, 79, 64, 0.08);
  border-radius: 18px;
  padding: 14px 16px;
  box-shadow: 0 8px 24px rgba(31,42,34,0.06);
}
.stat-label { text-transform: uppercase; letter-spacing: 0.08em; font-size: 0.72rem; color: #71675a; }
.stat-value { font-size: 1.8rem; font-weight: 700; color: #1f4f40; margin-top: 6px; }

.callout {
  background: #fffdf8;
  border-left: 5px solid #8e5a2b;
  padding: 14px 18px;
  border-radius: 10px;
  margin: 18px 0;
}
@media (max-width: 900px) { .stat-grid { grid-template-columns: repeat(2, minmax(0, 1fr)); } }
@media (max-width: 560px) { .stat-grid { grid-template-columns: 1fr; } }
</style>

<div class="hero">
  <p>A long-term study of breeding birds in nest boxes on and around the University of Southern Denmark campus. The project tracks breeding timing, occupancy, eggs, chicks, and year-to-year ecological variation in a simple woodland box system.</p>
</div>
```

## At a glance

```{r stats-cards}
#| results: asis
cat(sprintf(
  '<div class="stat-grid">
    <div class="stat-card"><div class="stat-label">Years monitored</div><div class="stat-value">%s</div></div>
    <div class="stat-card"><div class="stat-label">Nest visits</div><div class="stat-value">%s</div></div>
    <div class="stat-card"><div class="stat-label">Box IDs</div><div class="stat-value">%s</div></div>
    <div class="stat-card"><div class="stat-label">Occupied nest-years</div><div class="stat-value">%s</div></div>
    <div class="stat-card"><div class="stat-label">Median clutch size</div><div class="stat-value">%s</div></div>
  </div>',
  years_monitored, comma(total_visits), total_box_ids,
  comma(total_occupied_nests), median_clutch_size
))
```

## About the project

The SDU Nest Box Project is built around a straightforward idea: by following the same nest boxes through spring and early summer, year after year, it becomes possible to see how breeding birds respond to changing seasonal conditions. The project records when birds begin building nests, when eggs appear, when chicks hatch, and how breeding progresses across the season.

Because the same box system is monitored repeatedly, the project provides a consistent long-term view of breeding phenology and nest-box use. It is useful both as a research platform and as a field-based teaching project involving students and volunteers.

## Study area

The study takes place in mixed woodland around the University of Southern Denmark campus in Odense. The nest boxes are spread across a compact and accessible landscape, which makes the site well suited to regular monitoring and repeated seasonal comparisons.

```{r study-map}
#| fig-width: 8.5
#| fig-height: 7.8
ggplot(study_area_summary, aes(x = longitude, y = latitude)) +
  geom_point(shape = 21, size = 4.2, fill = "#d8c07a", colour = "white", stroke = 0.6) +
  geom_text(aes(label = box_number), size = 2.2, colour = "#203228") +
  coord_equal() +
  labs(
    title    = "Nest box locations in the study area",
    subtitle = "Each point is a monitored nest-box location",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text = element_blank(), panel.grid.minor = element_blank())
```

## What we record

During nest checks, the project records a simple set of field observations that are biologically informative and easy to compare across years:

- whether a box is empty or in use
- nest-building stage
- eggs and chicks
- species where known
- timing of breeding events
- broad breeding outcomes

Together, these observations make it possible to describe both the progress of a single breeding season and broader long-term trends.

## Current season snapshot

```{r current-season-text}
#| results: asis
if (!has_api) {
  cat('<div class="callout">Current-season data not available.</div>')
} else {
  cat(sprintf(
    '<div class="callout">The latest current-season data are for <strong>%s</strong>, updated to <strong>%s</strong>. So far there have been <strong>%s</strong> recorded checks across <strong>%s</strong> boxes, with <strong>%s</strong> boxes containing eggs and <strong>%s</strong> containing chicks.</div>',
    current_year, current_summary$latest_date, current_summary$checks,
    current_summary$boxes_checked, current_summary$boxes_with_eggs,
    current_summary$boxes_with_chicks
  ))
}
```

```{r current-season-plot}
#| fig-width: 9
#| fig-height: 4.5
#| eval: !expr has_api
ggplot(checks_by_day, aes(x = record_date, y = n_checks)) +
  geom_col(fill = "#214f3f", width = 0.8) +
  geom_text(aes(label = n_checks), vjust = -0.35, size = 3.8) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = paste("Checks recorded in", current_year),
    subtitle = "Number of nest checks recorded per day this season",
    x = NULL, y = "Checks"
  ) +
  theme_minimal(base_size = 13)
```

## Long-term results

### Occupancy by year

```{r occupancy-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(occupancy_by_year, aes(x = year, y = occupied_nests)) +
  geom_col(fill = "#9a5f2d", width = 0.72) +
  geom_line(aes(y = nests_with_chicks), colour = "#2f6b4f", linewidth = 1.1) +
  geom_point(aes(y = nests_with_chicks), colour = "#2f6b4f", size = 2) +
  labs(
    title    = "Occupied nests by year",
    subtitle = "Bars show occupied nest-years; the line shows nest-years with chicks",
    x = NULL, y = "Nest-years"
  ) +
  theme_minimal(base_size = 13)
```

### Timing of first eggs

```{r first-egg-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(first_egg_by_year, aes(x = year, y = median_first_egg_day)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#d8c07a", alpha = 0.45) +
  geom_line(colour = "#7a1f5c", linewidth = 1.1) +
  geom_point(colour = "#7a1f5c", size = 2.2) +
  labs(
    title    = "Median timing of first eggs by year",
    subtitle = "The ribbon shows the interquartile range around the median",
    x = NULL, y = "Estimated first egg day of year"
  ) +
  theme_minimal(base_size = 13)
```

### Species using the boxes

```{r species-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(species_summary, aes(x = reorder(species_public, n), y = n, fill = species_public)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Species recorded in occupied nest-years",
    subtitle = "Great Tits and Blue Tits dominate the system",
    x = NULL, y = "Occupied nest-years"
  ) +
  theme_minimal(base_size = 13)
```

## Why it matters

Long-term monitoring makes it possible to see that breeding timing and occupancy vary substantially between years. Those differences help show how local bird populations respond to changing spring conditions, habitat, and weather. Because the same nest-box system is monitored repeatedly, the project provides a rare combination of continuity, simplicity, and ecological value.

The project also matters as a teaching and engagement platform. It gives students field experience, supports practical training in ecology, and creates a structured way for volunteers to contribute to biodiversity monitoring.

## People and participation

This project has involved many students and volunteers over the years. If you are interested in contributing — whether as a student, researcher, or volunteer — please get in touch via the [Contact](contact.qmd) page.

## Animal welfare and ethics

Nest-box monitoring is carried out with care to minimise disturbance. Any ringing activities are conducted under the relevant permits and in accordance with Danish animal welfare legislation.
