---
title: "SDU Nest Box Project"
date-modified: last-modified
execute:
  freeze: auto
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
library(tibble)

# Standalone summary data for the public-facing site.
# These values are a static snapshot derived from the SDUBirdProject data.
# Update them manually here when the public page needs refreshing.
project_summary <- tibble(
  metric = c("Years monitored", "Box IDs recorded", "Nest visits", "Occupancy peak", "Median first egg"),
  value  = c("12", "200", "19,487", "86 nests", "Day 112")
)

occupancy_by_year <- tibble(
  year = c(2013, 2014, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  occupied_nests = c(71, 86, 76, 72, 65, 55, 72, 59, 69, 69, 61, 63),
  nests_with_chicks = c(69, 83, 52, 53, 50, 51, 65, 37, 46, 60, 54, 51)
)

timing_by_year <- tibble(
  year = c(2013, 2014, 2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024, 2025),
  median_first_egg_day = c(122, 109, 113, 107, 113, 108, 110, 116, 112, 112, 111, 106),
  q25 = c(119, 102, 108, 104, 111, 101, 107, 113, 110, 109.5, 108, 103),
  q75 = c(126, 112, 120, 114, 118, 112.5, 114, 118, 116.5, 115, 118.25, 109)
)

species_summary <- tibble(
  species = c("Great Tit", "Blue Tit", "Marsh Tit", "Nuthatch", "Other"),
  nest_years = c(528, 203, 12, 10, 1)
)
```

```{=html}
<style>
.nest-hero {
  background: linear-gradient(135deg, #204f40 0%, #826035 100%);
  color: #fbfaf6;
  padding: 30px 34px;
  border-radius: 24px;
  margin: 0 0 24px 0;
  box-shadow: 0 16px 36px rgba(31, 79, 64, 0.18);
}
.nest-hero p {
  margin: 0;
  max-width: 860px;
  color: rgba(251, 250, 246, 0.92);
  font-size: 1.05rem;
}
.nest-stat-grid {
  display: grid;
  grid-template-columns: repeat(5, minmax(0, 1fr));
  gap: 14px;
  margin: 22px 0 28px 0;
}
.nest-stat-card {
  background: #fbfaf7;
  border: 1px solid rgba(32, 79, 64, 0.08);
  border-radius: 18px;
  padding: 14px 16px;
  box-shadow: 0 8px 24px rgba(31, 42, 34, 0.06);
}
.nest-stat-label {
  text-transform: uppercase;
  letter-spacing: 0.08em;
  font-size: 0.72rem;
  color: #71675a;
}
.nest-stat-value {
  font-size: 1.75rem;
  font-weight: 700;
  color: #204f40;
  margin-top: 6px;
}
.nest-photo-grid {
  display: grid;
  grid-template-columns: repeat(2, minmax(0, 1fr));
  gap: 18px;
  margin: 16px 0 28px 0;
}
.nest-map-frame {
  margin: 14px 0 10px 0;
  border: 1px solid var(--border-soft, #dbe2ea);
  border-radius: 14px;
  overflow: hidden;
  box-shadow: 0 8px 24px rgba(31, 42, 34, 0.06);
}
.nest-map-frame iframe {
  width: 100%;
  height: 420px;
  border: 0;
  display: block;
}
.nest-map-note {
  color: #5a5a5a;
  font-size: 0.95rem;
  margin-top: 0.5rem;
}
.nest-photo-placeholder {
  min-height: 260px;
  border: 1px dashed #9aaec1;
  border-radius: 14px;
  background: #f7f9fb;
  color: #5a5a5a;
  display: flex;
  align-items: center;
  justify-content: center;
  text-align: center;
  padding: 1rem;
}
@media (max-width: 900px) {
  .nest-stat-grid {
    grid-template-columns: repeat(2, minmax(0, 1fr));
  }
  .nest-photo-grid {
    grid-template-columns: 1fr;
  }
}
@media (max-width: 560px) {
  .nest-stat-grid {
    grid-template-columns: 1fr;
  }
}
</style>
```

```{=html}
<div class="nest-hero">
  <p>The SDU Nest Box Project is a long-term field study of breeding birds in nest boxes on and around the University of Southern Denmark campus. It combines simple repeated field observations with long-term ecological questions about breeding phenology, occupancy, reproductive output, and year-to-year environmental variation.</p>
</div>
```

## At a glance

```{r stats-cards}
#| results: asis
cat(
  '<div class="nest-stat-grid">',
  paste0(
    '<div class="nest-stat-card"><div class="nest-stat-label">',
    project_summary$metric,
    '</div><div class="nest-stat-value">',
    project_summary$value,
    '</div></div>',
    collapse = ""
  ),
  '</div>'
)
```

## About the project

The project follows nest boxes through spring and early summer to record when birds begin breeding, how many nests are occupied, when eggs appear, when chicks hatch, and how breeding progresses across the season. Because the same box system has been monitored across many years, the project provides a compact but informative long-term record of breeding ecology in a local woodland landscape.

This work matters for both research and teaching. It supports field-based training for students, provides a platform for final-year and undergraduate research projects, and creates a long-term dataset that can be used to ask ecological questions about variation in breeding timing and reproductive performance.

## Study area map

The project is based around the woodland and green areas on and around the SDU campus in Odense. The map below shows the broader study area rather than precise nest-box locations.

```{=html}
<div class="nest-map-frame">
  <iframe
    title="Map of the SDU Odense study area"
    loading="lazy"
    src="https://www.openstreetmap.org/export/embed.html?bbox=10.409%2C55.360%2C10.439%2C55.377&layer=mapnik&marker=55.3682%2C10.4240">
  </iframe>
</div>
<p class="nest-map-note">
  Approximate study area centred on SDU Odense. <a href="https://www.openstreetmap.org/?mlat=55.3682&amp;mlon=10.4240#map=15/55.3682/10.4240">Open larger map</a>.
</p>
```

## Photos

```{=html}
<div class="nest-photo-grid">
  <div class="nest-photo-placeholder">
    Add project photo here<br><code>images/nest-box-project/fieldwork.jpg</code>
  </div>
  <div class="nest-photo-placeholder">
    Add project photo here<br><code>images/nest-box-project/nest-boxes.jpg</code>
  </div>
</div>
```

## What the project records

Typical field observations include:

- whether a nest box is empty or occupied
- nest-building stage
- eggs and chicks
- species identity where known
- timing of breeding events
- broad breeding outcomes

Together, these observations make it possible to describe both single breeding seasons and broader long-term changes in phenology and nest-box use.

## Long-term summaries

### Occupancy by year

```{r occupancy-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(occupancy_by_year, aes(x = year)) +
  geom_col(aes(y = occupied_nests), fill = "#8d5a2b", width = 0.7) +
  geom_line(aes(y = nests_with_chicks), colour = "#2f6b4f", linewidth = 1.1) +
  geom_point(aes(y = nests_with_chicks), colour = "#2f6b4f", size = 2) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Occupied nests by year",
    subtitle = "Bars show occupied nest-years; the line shows nest-years with chicks",
    x = NULL,
    y = "Nest-years"
  ) +
  theme_minimal(base_size = 13)
```

### Timing of first eggs

```{r first-egg-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(timing_by_year, aes(x = year, y = median_first_egg_day)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#d8c07a", alpha = 0.45) +
  geom_line(colour = "#7a1f5c", linewidth = 1.1) +
  geom_point(colour = "#7a1f5c", size = 2.2) +
  labs(
    title = "Median timing of first eggs by year",
    subtitle = "The ribbon shows the interquartile range around the median",
    x = NULL,
    y = "Estimated first egg day of year"
  ) +
  theme_minimal(base_size = 13)
```

### Species using the boxes

```{r species-plot}
#| fig-width: 8.5
#| fig-height: 4.8
ggplot(species_summary, aes(x = reorder(species, nest_years), y = nest_years, fill = species)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Species recorded in occupied nest-years",
    subtitle = "Great Tits and Blue Tits dominate the system",
    x = NULL,
    y = "Occupied nest-years"
  ) +
  theme_minimal(base_size = 13)
```

## Research outputs

The nest box project has supported ecological field research, student-led projects, and teaching-led data collection. This section can be expanded over time as outputs accumulate.

### Selected papers

- Corregidor-Castro, A. and Jones, O. R. (2021). The effect of nest temperature on growth and survival in juvenile Great Tits *Parus major*. *Ecology and Evolution*, 11, 7346-7353. [DOI](https://doi.org/10.1002/ece3.7565)  
  Experimental warming of nests showed that relatively small temperature increases can reduce nestling size and mass, linking early developmental conditions to later performance under climate change.
- Add journal article citation here.
- Add methods or field-note publication here.

### Student project areas

Student research linked to the project has included work on:

- phenology of breeding
- occupancy and nest preferences
- foraging behaviour
- cognition

## Why it matters

Long-term monitoring of a simple nest-box system makes it possible to see how breeding timing, occupancy, and reproductive outcomes vary between years. That continuity gives the project value well beyond a single field season: it supports ecological inference, student training, and public-facing biodiversity engagement from the same core dataset.

## Participation and contact

The project has involved many students and volunteers over time. If you are interested in the project as a student, collaborator, or volunteer, please get in touch through the [Contact](contact.qmd) page.

## Animal welfare and ethics

Nest-box monitoring is carried out with care to minimise disturbance. Any ringing activities are conducted under the relevant permits and in accordance with Danish animal welfare legislation.
